INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-04 chapters/language_definition.tex
finished introduction
tweak code listing

\section{Creating Language Definition in MPS}

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\textbf{grammar} \textit{SimpleXML};
	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
	             ;
	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
	             |   \antlrparserrule{element}
	             |   \antlrparserrule{comment}
	             |   \antlrlexerrule{CDATA}
	             ;
	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
	             |   \antlrlexerrule{DIGIT}
	             ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
\end{antlr}
\caption{SimpleXML grammar}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
             |   \antlrlexerrule{DIGIT}
             ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
\end{antlr}
\caption{Lexer rule from SimpleXML}
\label{fig:lexerrule}
\end{figure*}

@@

% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
\begin{figure*}[ht]
\centering
\begin{antlr}
	Flatten(\textit{R}):
	1) Define \textit{T} as a list of empty string
	2) For each alternative \textit{A} of the rule \textit{R}:
	3)     Define \textit{R} = \antlrap\antlrap
	4)     For each element \textit{E} of \textit{A}:
	5)         If \textit{E} is not yet flattened:
	6)             Flatten(\textit{E})
	7)         \textit{R}.append(\textit{E})
	8)         \textit{R}.append(\textit{E}.operator)
	9)     \textit{T}.add(\textit{R})
	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
	12) Return \textit{S}
\end{antlr}
\caption{Flatten algorithm}
\label{fig:algflatten}
\end{figure*}

@@

However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
The structure is created according to object-oriented principles.
Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
             \textcolor{gray}{;}
\end{antlr}
\caption{Fragment of the SimpleXML grammar}
\label{fig:xmlgrammarelement}
\end{figure*}

Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
           |   \antlrparserrule{element}
           |   \antlrparserrule{comment}
           |   \antlrlexerrule{CDATA}
           ;
\end{antlr}
\caption{Content rule}
\label{fig:contentrule}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\begin{mps}
	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
           |   \mpsconcept{Content{\_}2}
           |   \mpsconcept{Content{\_}3}
           |   \mpsconcept{Content{\_}4}
\end{mps}
\caption{IContent interface}
\label{fig:icontentitf}
\end{figure*}

References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:elementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
           ;
\end{antlr}
\caption{Element rule}
\label{fig:elementrule}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
           |   \antlrparserrule{element}
           |   \antlrparserrule{comment}
           |   \antlrlexerrule{CDATA}
           ;

	\antlrparserrule{element}    :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
           |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
           ;

	\antlrparserrule{comment}    :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
\end{antlr}
\caption{Content parser rule with children}
\label{fig:contentrulewithchildren}
\end{figure*}

@@

% TODO pouzit standardni environment pro algoritmy
\begin{figure*}[ht]
\centering
\begin{antlr}
	FindPathsToEndNodes($R$):
	1)  Define $L$ as an empty list of list of nodes
	2)  Return FindPathsToEndNodes($R$, $L$)

	FindPathsToEndNodes($R$, $L$):
	3)  Define $Q$ as list of list of nodes
	4)  For each alternative $A$ of rule $R$:
	5)      $L1$ = Clone($L$)
	6)      If $A$ is a parser rule with only one element $E$:
	7)          Let $I$ be interface/concept representing rule $E$
	8)          $L1$.Add($I$)
	9)          $P$ = FindPathsToEndNodes($E$, $L1$)
	10)         $Q$ = Merge($Q$, $P$)
	11)     Else
	12)         $L1$.Add($R$)
	13)         Let $P$ be concept representing $A$
	14)         $L1$.Add($P$)
	15)         $Q$.Add($L1$)
	16) Return $Q$
\end{antlr}
\caption{Algorithm for finding end rules}
\label{fig:shortcut_algorithm}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
	\antlrlexerrule{STRING}      :   \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
	            |   \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'}
	            ;
\end{antlr}
\caption{Definition of an XML attribute}
\label{fig:xmlattribute}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{attribute}   :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
	            |   \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'}
	            ;
	\antlrlexerrule{TEXT1}       :   \antlrregex{~["]*} ;
	\antlrlexerrule{TEXT2}       :   \antlrregex{~[']*} ;
\end{antlr}
\caption{Adjusted grammar of SimpleXML}
\label{fig:xmladjustgrammar}
\end{figure*}

@@

As the second example, we will show the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{eos}            : \antlrlexerrule{SemiColon}
	               | \antlrlexerrule{EOF}
	               | {lineTerminatorAhead()}?
	               | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}?
	               ;
	\textcolor{gray}{// Example reference of the eos rule}
	\antlrparserrule{breakStatement} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos}
	               ;
\end{antlr}
\caption{Statements in JavaScript}
\label{fig:javascriptstmt}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{eos} : \antlrliteral{;} ;
\end{antlr}
\caption{eos parser rule}
\label{fig:eosrule}
\end{figure*}

>>>>

\section{Creating Language Definition in MPS}
\label{sect:LANGDEF}

@@

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	\textbf{grammar} \textit{SimpleXML};
	\antlrparserrule{document}  : \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
	\antlrparserrule{prolog}    : \antlrliteral{<?xml } \antlrparserrule{attrib}* \antlrliteral{?>} ;
	\antlrparserrule{comment}   : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
	\antlrparserrule{element}   : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
	\antlrparserrule{attrib}    : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
	\antlrparserrule{content}   : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
	\antlrlexerrule{Name}      : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}     : \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}  : \antlrlexerrule{NameStartChar} | \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
	\antlrlexerrule{TEXT}      : \antlrregex{~[<"]*} ;
	\antlrlexerrule{CDATA}     : \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
\end{alltt}
\end{framed}
\caption{SimpleXML grammar}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrlexerrule{Name}     : \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}    : \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar} : \antlrlexerrule{NameStartChar} 
	         | \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.} | \antlrlexerrule{DIGIT} ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar} : \antlrregex{[:a-zA-Z]} ;
\end{alltt}
\end{framed}
\caption{Lexer rule from SimpleXML}
\label{fig:lexerrule}
\end{figure}

@@

% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
Flatten(\textit{R}):
  Define \textit{T} as a list of empty string
  For each alternative \textit{A} of the rule \textit{R}:
    Define \textit{R} = \antlrap\antlrap
    For each element \textit{E} of \textit{A}:
	  If \textit{E} is not yet flattened:
	    Flatten(\textit{E})
      \textit{R}.append(\textit{E})
      \textit{R}.append(\textit{E}.operator)
    \textit{T}.add(\textit{R})
  Build a string S from elements of \textit{T}:
    (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
  Return \textit{S}
\end{alltt}
\end{framed}
\caption{Flatten algorithm}
\label{fig:algflatten}
\end{figure}

@@

However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
The structure is created according to object-oriented principles.
Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
For example, consider the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlelementrule}.
Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrap<\antlrap Name attrib* \antlrap/>\antlrap ;
\end{alltt}
\end{framed}
\caption{Element rule from the SimpleXML grammar}
\label{fig:xmlelementrule}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{content}    :   \antlrlexerrule{TEXT}
           |   \antlrparserrule{element}
           |   \antlrparserrule{comment}
           |   \antlrlexerrule{CDATA}
           ;
\end{alltt}
\end{framed}
\caption{Content rule}
\label{fig:contentrule}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\mpsinterface{IContent}   :   \mpsconcept{Content{\_}1}
           |   \mpsconcept{Content{\_}2}
           |   \mpsconcept{Content{\_}3}
           |   \mpsconcept{Content{\_}4}
\end{alltt}
\end{framed}
\caption{IContent interface}
\label{fig:icontentitf}
\end{figure}

References to other parser rules are captured by links to children (interfaces or object nodes) in the AST nodes.
For the \antlrparserrule{element} rule that is referencing the \antlrparserrule{content} rule, defined as in Figure~\ref{fig:xmlelementrule}, we get the AST node representing the first alternative that is shown in Figure~\ref{fig:element_node_full}.
There is one \mpsinterface{IElement} interface node that is implemented by two object nodes \mpsconcept{Element{\_}1}, \mpsconcept{Element{\_}2}.
The node \mpsconcept{Element{\_}1} has two child links pointing to \mpsinterface{IAttribute} and \mpsinterface{IContent}.

@@

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{content} : \antlrlexerrule{TEXT} | \antlrparserrule{element} | \antlrparserrule{comment} | \antlrlexerrule{CDATA} ;
	\antlrparserrule{element} : \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>} | \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attrib}* \antlrliteral{/>} ;
	\antlrparserrule{comment} : \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
\end{alltt}
\end{framed}
\caption{Content parser rule with children}
\label{fig:contentrulewithchildren}
\end{figure*}

@@

% TODO pouzit standardni environment pro algoritmy
\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	FindPathsToEndNodes($R$):
	  Define $L$ as an empty list of list of nodes
	  Return FindPathsToEndNodes($R$, $L$)

	FindPathsToEndNodes($R$, $L$):
	  Define $Q$ as list of list of nodes
	  For each alternative $A$ of rule $R$:
	    $L1$ = Clone($L$)
	    If $A$ is a parser rule with only one element $E$:
	      Let $I$ be interface/concept representing rule $E$
	      $L1$.Add($I$)
	      $P$ = FindPathsToEndNodes($E$, $L1$)
	      $Q$ = Merge($Q$, $P$)
	    Else
	      $L1$.Add($R$)
	      Let $P$ be concept representing $A$
	      $L1$.Add($P$)
	      $Q$.Add($L1$)
	  Return $Q$
\end{alltt}
\end{framed}
\caption{Algorithm for finding end rules}
\label{fig:shortcut_algorithm}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
	\antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
	       | \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'} ;
\end{alltt}
\end{framed}
\caption{Definition of an XML attribute}
\label{fig:xmlattribute}
\end{figure}

@@

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
	       | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
	\antlrlexerrule{TEXT1} : \antlrregex{~["]*} ;
	\antlrlexerrule{TEXT2} : \antlrregex{~[']*} ;
\end{alltt}
\end{framed}
\caption{Adjusted grammar of SimpleXML}
\label{fig:xmladjustgrammar}
\end{figure}

@@

As the second example, we will use the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineTerminatorAhead()}? 
	    | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}? ;
	\textcolor{gray}{// Example reference of the eos rule}
	\antlrparserrule{breakStmt} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos} ;
\end{alltt}
\end{framed}
\caption{Statements in JavaScript}
\label{fig:javascriptstmt}
\end{figure*}

@@

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{eos} : \antlrliteral{;} ;
\end{alltt}
\end{framed}
\caption{eos parser rule}
\label{fig:eosrule}
\end{figure}
